### 2020-01-28 Mathematical Statistics/05_conf_int_bootstrap.rmd
pre/post reconstruct
pre/post reconstruct
bootstrap work
bootstrap work

library(ggplot2, quietly = TRUE, warn.conflicts = FALSE)
library(ggrepel, quietly = TRUE, warn.conflicts = FALSE)
library(ggthemes, quietly = TRUE, warn.conflicts = FALSE)
library(knitr, quietly = TRUE, warn.conflicts = FALSE)
library(kableExtra, quietly = TRUE, warn.conflicts = FALSE)
library(Rblpapi, quietly = TRUE, warn.conflicts = FALSE)
library(scales, quietly = TRUE, warn.conflicts = FALSE)
library(pander, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(formattable, quietly = TRUE, warn.conflicts = FALSE)
library(grid, quietly = TRUE, warn.conflicts = FALSE)
library(gridExtra, quietly = TRUE, warn.conflicts = FALSE)
library(png, quietly = TRUE, warn.conflicts = FALSE)
library(extrafont, quietly = TRUE, warn.conflicts = FALSE)
library(tinytex, quietly = TRUE, warn.conflicts = FALSE)
library(stringr, quietly = TRUE, warn.conflicts = FALSE)
library(lubridate, quietly = TRUE, warn.conflicts = FALSE)
library(reshape2, quietly = TRUE, warn.conflicts = FALSE)
library(ggrepel, quietly = TRUE, warn.conflicts = FALSE)
library(gtools, quietly = TRUE, warn.conflicts = FALSE)
library(here, quietly = T, warn.conflicts = F)
library(boot, quietly = T, warn.conflicts = F)
library(ggExtra, quietly = T, warn.conflicts = F)

@@

```{r}
combinations(n = 3, r = 4, v = samples, repeats.allowed = T)
```

@@

```{r}
Bangladesh <- data.table(read.csv(paste0(data.dir, "Bangladesh.csv"),
                                 header = T))

head(Bangladesh)

@@

>>>>

library(ggplot2, quietly = TRUE, warn.conflicts = FALSE)
library(ggrepel, quietly = TRUE, warn.conflicts = FALSE)
library(ggthemes, quietly = TRUE, warn.conflicts = FALSE)
library(knitr, quietly = TRUE, warn.conflicts = FALSE)
library(kableExtra, quietly = TRUE, warn.conflicts = FALSE)
library(scales, quietly = TRUE, warn.conflicts = FALSE)
library(pander, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(formattable, quietly = TRUE, warn.conflicts = FALSE)
library(grid, quietly = TRUE, warn.conflicts = FALSE)
library(gridExtra, quietly = TRUE, warn.conflicts = FALSE)
library(png, quietly = TRUE, warn.conflicts = FALSE)
library(extrafont, quietly = TRUE, warn.conflicts = FALSE)
library(tinytex, quietly = TRUE, warn.conflicts = FALSE)
library(stringr, quietly = TRUE, warn.conflicts = FALSE)
library(lubridate, quietly = TRUE, warn.conflicts = FALSE)
library(reshape2, quietly = TRUE, warn.conflicts = FALSE)
library(ggrepel, quietly = TRUE, warn.conflicts = FALSE)
library(gtools, quietly = TRUE, warn.conflicts = FALSE)
library(here, quietly = T, warn.conflicts = F)
library(boot, quietly = T, warn.conflicts = F)
library(ggExtra, quietly = T, warn.conflicts = F)

@@

```{r}
combinations(n = 3, r = 4, v = samples, repeats.allowed = T)

choose(4 + 3 - 1, 4)
```

@@

```{r}
Bangladesh <- data.table(read.csv(paste0(data.dir, "Bangladesh.csv"),
                                 header = T))
head(Bangladesh)

@@

### 5.12

Consider Bangladesh chlorine (concentration). Bootstrap the trimmed mean (say, trim the upper and lower 25%), and compare your results with the usual mean (previous result).

```{r}
values <- Bangladesh[!is.na(Chlorine)]$Chlorine

n <- length(values); N <- 10e3; trim <- .25

observed <- mean(values, trim = trim)
bootstrap <- vector(mode = "numeric", length = N)

for(i in 1:N)
{
   bootstrap[i] <- mean( sample(values, n, replace = T), trim = trim )
}

ggplot(data.table(value = bootstrap), aes(value)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(col = "darkorange")

alpha <- 0.05

quantile(bootstrap, c(alpha/2, 1 - alpha/2))

# boot pkg

boot.fn <- function(data, index){
   mean( data[index], trim = trim)
}

boot(values, boot.fn, R = N)
```

The bootstrap 20% trimmed mean is substantially smaller than the regular mean. The confidence intervals are also tighter, which is unsuprising due to the heavy presence of outliers in our sample.

### 5.13

The data set _FishMercury_ contains mercury levels (parts per million) for 30 fish caught in lakes in Minnesota.

```{r}
FishMercury <- data.table(read.csv(paste0(data.dir, "FishMercury.csv"),
                                 header = T))
```

a.) Create a histogram or boxplot of the data. What do you observe?

```{r}
ggplot(FishMercury, aes(Mercury)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(col = "darkorange") +
   geom_rug(col = "darkred")

FishMercury[Mercury > 1.5]

ggplot(FishMercury[Mercury < 1.5], aes(Mercury)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(col = "darkorange") +
   geom_rug(col = "darkred")

FishMercury[Mercury > 1.5] / sd(FishMercury$Mercury) # 6 sd outlier

One extreme (6 SD) outlier in the data.

b.) Bootstrap the mean, and record the bootstrap standard error and the 95% bootstrap percentile interval.

```{r}
n <- nrow(FishMercury); N <- 10e3

observed <- mean(FishMercury$Mercury)

bootstrap <- vector(mode = "numeric", length = n)

for(i in 1:N)
{
   bootstrap[i] <- mean( sample(FishMercury$Mercury, n, replace = T) )
}

alpha <- 0.05

quantile(bootstrap, c(alpha/2, 1 - alpha/2))

# boot pkg

boot.fn <- function(data, index) {
   mean( data[index] )
}

boot(FishMercury$Mercury, boot.fn, R = N)
```

c.) Remove the outlier and bootstrap the mean of the remaining data. Record the bootstrap standard error and the 95% bootstrap percentile interval.

```{r}
mercury <- FishMercury[Mercury < 1.5]$Mercury

n <- length(mercury)
bootstrap.values <- vector(mode = "numeric", length = n)

observed <- mean(mercury)

for( i in 1:N)
{
   bootstrap.values[i] <- mean( sample(mercury, size = n, replace = T) )   
}

mean(bootstrap.values)

boot(mercury, boot.fn, R = N)

quantile(bootstrap.values, c(alpha/2, 1 - alpha/2))
```

The standard error reduced drastically, and the confidence intervals are much more narrow.

### 5.14

In Section 3.3, we performed a permutation test to determine if men and women consumed, on average, different amounts of hot wings.

```{r}
BeerWings <- data.table(read.csv(paste0(data.dir, "Beerwings.csv"),
                                 header = T))

wings <- BeerWings$Hotwings

?stat_ecdf

ggplot(BeerWings, aes(Hotwings, group = Gender, col = Gender)) +
   stat_ecdf(geom = "step")

N <- 10e3; alpha <- 0.05; n <- length(wings)

wings.m <- BeerWings[Gender == "M"]$Hotwings
wings.f <- BeerWings[Gender == "F"]$Hotwings

observed <- mean(wings.m) - mean(wings.f)

permutation.values <- vector(mode = "numeric", length = N)

for(i in 1:N)
{
   samp <- sample(1:n, length(wings.m), replace = F)

   samp.m <- wings[samp]; samp.w <- wings[-samp]

   permutation.values[i] <- mean(samp.m) - mean(samp.w)
}

ggplot(data.table(values = permutation.values), aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(col = "darkorange") +
   geom_vline(xintercept = observed, col = "darkred")

ggplot(data.table(value = permutation.values), aes(value)) +
   stat_ecdf(geom = "step") +
   geom_vline(xintercept = observed, col = "darkred") +
   scale_y_continuous(labels = comma)

p <- (sum(permutation.values >= observed) + 1) / (N + 1)

var <- p*(1 - p)/N
```

a.) Bootstrap the difference in means and describe the bootstrap distribution.

```{r}
bootstrap.values <- vector(mode = "numeric", length = N)

for(i in 1:N)
{
   samp.m <- sample(wings.m, size = length(wings.m), replace = T)
   samp.f <- sample(wings.f, size = length(wings.f), replace = T)

   bootstrap.values[i] <- mean(samp.m) - mean(samp.f)
}

ggplot(data.table(values = bootstrap.values), aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30)

boot.mean <- mean(bootstrap.values)
boot.se <- sd(bootstrap.values)

boot.bias <- observed - boot.mean

boot.se; boot.bias
```

b.) Find a 95% bootstrap percentile confidence interval for the difference of means, and give a sentence interpreting this interval.

```{r}
quantile(bootstrap.values, c(alpha/2, 1 - alpha/2))
```

The 95% confidence interval does not contain 0, further supporting our permutation test results.

c.) How do the bootstrap and permutation distributions differ?

The permutation distribution is sampled without replacement, and the bootstrap distribution is sampled using replacement.

### 5.15

A high school student was curious about the total number of minutes devoted to commercials during any given half-hour time period on basic and extended cable TV channels. (B. Rodgers and T. Robinson, personal communication).

Import the TV data.

```{r}
TV <- data.table(read.csv(paste0(data.dir, "TV.csv"),
                                 header = T))
```

a.) Perform some exploratory data analysis and obtain summary statistics on the commercial times on basic and extended cable TV channels (do seperate analysis for each type of channel).

```{r}
ggplot(TV, aes(Cable, Times, fill = Cable)) +
   geom_boxplot()

ggplot(TV, aes(Times, group = Cable)) +
   geom_density(aes(y = ..density.., fill = Cable), alpha = .45)

aov(TV$Times ~ TV$Cable)

tv.basic <- TV[Cable == "Basic"]$Times
tv.extended <- TV[Cable == "Extended"]$Times
tv.pooled <- c(tv.basic, tv.extended)

observed <- mean(tv.basic) - mean(tv.extended)

mean(tv.basic); mean(tv.extended)
```

b.) Bootstrap the difference in mean times, plot the distribution, and give summary statistics of the bootstrap distribution. Obtain a 95% bootstrap percentile confidence interval, and interpret this interval.

```{r}
N <- 10e3; n <- length(TV$Times); alpha <- 0.05

bootstrap.values <- vector(mode = "numeric", N)

for(i in 1:N)
{
   samp.basic <- sample(tv.basic, size = length(tv.basic), replace = T)
   samp.extended <- sample(tv.extended, size = length(tv.extended), replace = T)

   bootstrap.values[i] <- mean(samp.basic) - mean(samp.extended)
}

boot.mean <- mean(bootstrap.values)
boot.se <- sd(bootstrap.values)
boot.bias <- observed - boot.mean

quantile(bootstrap.values, c(alpha/2, 1 - alpha/2)) # bootstrap interval
```

The 95% boostrap interval does not contain 0, suggesting there is in fact a difference in the times between basic and extended (basic being longer).

c.) What is the bootstrap estimate of the bias?

```{r}
boot.bias

boot.bias / boot.se # about 1% of the standard error
```

d.) Conduct a permutation test to see if the difference in mean commercial times is statistically significant, and state your conclusion.

```{r}
permutation.values <- vector(mode = "numeric", n = N)

for(i in 1:N)
{
   samp <- sample(length(tv.pooled), length(tv.basic), replace = F)

   samp.basic <- tv.pooled[samp]; samp.extended <- tv.pooled[-samp]

   permutation.values[i] <- mean(samp.basic) - mean(samp.extended)
}

ggplot(data.table(values = permutation.values), aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_vline(xintercept = observed, col = "darkred")

p <- ( sum(permutation.values >= observed) + 1) / ( N + 1)
var <- p*(1 - p)/N
```

Permutation test supports the alternative hypothesis that there is a difference in commercial times.
